### 2022-06-09 manuscript.tex
Merge pull request #225 from SORTEE-Github-Hackathon/RCO-software-to-code [ci skip]
This build is based on
https://github.com/SORTEE-Github-Hackathon/manuscript/commit/f4614eb536313e122f922ad46d0e4f9f246b724a.

This commit was created by the following CI build and job:
https://github.com/SORTEE-Github-Hackathon/manuscript/commit/f4614eb536313e122f922ad46d0e4f9f246b724a/checks
https://github.com/SORTEE-Github-Hackathon/manuscript/actions/runs/2466714947

This manuscript
(\href{https://SORTEE-Github-Hackathon.github.io/manuscript/v/55dfe2ab70354dd7be425790d6bcbeb459fbb92e/}{permalink})
was automatically generated
from \href{https://github.com/SORTEE-Github-Hackathon/manuscript/tree/55dfe2ab70354dd7be425790d6bcbeb459fbb92e}{SORTEE-Github-Hackathon/manuscript@55dfe2a}
on June 9, 2022.

@@

Most scientists, including ecologists and evolutionary biologists, are increasingly dependent on computational tools in their research\textsuperscript{\protect\hyperlink{ref-fJWFe93e}{1}}.
Researchers write and use code as part of their scientific workflow to perform a wide-variety of tasks ranging from data management, data analysis, and study replication, to the application and the development of tools for hypothesis testing.
To maintain code for scientific collaboration requires an efficient and well-documented work-flow\textsuperscript{\protect\hyperlink{ref-1Kqna6l2}{2}}.
To facilitate this process, scientists have been increasingly adopting tools from information and systems technology, such as cloud-based services for documentation and version control (\emph{e.g.}, from the Google Suite, with Docs, Sheets, and Drive; the Microsoft Suite, with Word, Excel, and OneDrive; and GitHub;\textsuperscript{\protect\hyperlink{ref-10ghgV3S8}{3}}).
However, most researchers lack exposure to adequate software development practices and are required to dedicate valuable time and effort to self-teach the use of research-facilitating tools, and thus may be limited in their ability to adhere to adequate standards of scientific code quality and maintenance\textsuperscript{\protect\hyperlink{ref-fJWFe93e}{1},\protect\hyperlink{ref-10SpoByIw}{4},\protect\hyperlink{ref-O6UbstGG}{5}}.
Here, we review and discuss one of the most used web-based platforms for computational version control and collaboration, GitHub, and provide researchers in ecology and evolutionary biology (EEB) with practical workflows aimed at facilitating their scientific code and management process.

With over 73 million registered users, as of 2022, GitHub is the most widely-used web platform for collaborating on computer code\textsuperscript{\protect\hyperlink{ref-nwCtHDCn}{6}}.
GitHub provides a simplified but powerful web interface that allows users to participate in projects, contribute code, report and discuss software bugs, discover existing code and data, and publish new code.
Through version control, users have a detailed, chronological record of the files and directories stored in their repositories (see \protect\hyperlink{definitions}{Box 1};\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7}}).
This workflow provides a strong and clear advantage over receiving, processing and sending files back-and-forth (e.g.~via email), a process that can easily become challenging and time-consuming in projects extending in time and in the number of collaborators\textsuperscript{\protect\hyperlink{ref-4ny1onB0}{8}}.
Through the combination of version control management and the network- and collaboration-based features, GitHub can broadly facilitate openly available source code alongside concomitant collaborative development\textsuperscript{\protect\hyperlink{ref-kEX5dgzK}{9}}.

Git is the version control software system that enables all the collaborative tools available on GitHub.
Although the understanding of basic concepts of Git (such as commit, push, pull, checkout; see \protect\hyperlink{definitions}{Box 1}) is necessary, the GitHub web-based platform and its integrated development environments (such as the GitHub Desktop) allow users to perform most repository and data management operations without opening Git command-line sessions.
Still, GitHub users are encouraged to explore and improve their proficiency in Git when feeling comfortable, so they can feel confident in performing more flexible, complex operations (such as integrating changes from one branch into another branch with \texttt{git\ merge} and \texttt{git\ rebase}). Examples of extensive explanations on Git can be found in journal articles\textsuperscript{\protect\hyperlink{ref-kEX5dgzK}{9},\protect\hyperlink{ref-PlcxShQU}{10}}, video tutorials\textsuperscript{\protect\hyperlink{ref-NIS0JOW0}{11}}, and books\textsuperscript{\protect\hyperlink{ref-13jOlVcpp}{12}}.

The expansive GitHub user-community and the numerous resources on how to use GitHub have boosted its growing popularity\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7},\protect\hyperlink{ref-kEX5dgzK}{9},\protect\hyperlink{ref-13jOlVcpp}{12}/,\protect\hyperlink{ref-u5aEVE4B}{13}}.
Nevertheless, although multiple articles have encouraged researchers in EEB to adopt GitHub as part of their research process\textsuperscript{\protect\hyperlink{ref-10ghgV3S8}{3},\protect\hyperlink{ref-3DKwn1sY}{14}}, its use is still not widespread.
First-time users without formal training in information technology may face steep learning curves because GitHub and its features have been centered on collaboration for software development in information systems\textsuperscript{\protect\hyperlink{ref-139b0pSGc}{15}}.
Moreover, domain-specific perspectives and resources providing tractable examples and practical guidance for researchers in EEB on GitHub are scarce (but see https://ourcodingclub.github.io; https://www.openscapes.org).
A common adoption of GitHub for collaborating on a variety of research tasks can ultimately enable EEB researchers to spend less time on creating novel processes for collaboration and more time on their research\textsuperscript{\protect\hyperlink{ref-ydrk01SR}{16}}.
More importantly, increasing the availability of data and code management standards -- of which GitHub is one increasingly important component -- make research more reproducible and collaborative\textsuperscript{\protect\hyperlink{ref-13QX8XU3J}{17}}.

@@

\textbf{release}: At any point a release can be made on GitHub to mark a significant milestone in the progression of a \textbf{repository}. While this GitHub feature is designed with releases of new versions of software in mind (\emph{e.g.}, v1.0.0), it can also be used to create a snapshot of a repository at significant stages like submission, revision, and acceptance of an associated manuscript. \\

@@

Peer review plays a critical role in the scientific evaluation of EEB research.
GitHub provides an open and transparent platform that can be used for either directly providing feedback on research products or building a to-do list from reviewer comments.
Peer review of research software by the open software development community rOpenSci (https://ropensci.org/software-review/) and of research software and associated manuscripts by the Journal of Open Source Software (https://joss.readthedocs.io/en/latest/submitting.html) requires that submitted work is hosted on GitHub and their review processes make use of GitHub issues.
rOpenSci's efforts have resulted in many well-used R packages for ecology research including rfishbase\textsuperscript{\protect\hyperlink{ref-Xsdcv6q}{33}} and taxize\textsuperscript{\protect\hyperlink{ref-FVBWKkZu}{34}}.

@@

Some specialized journals, such as the the Journal of Open Source Software (\url{https://joss.readthedocs.io/en/latest/submitting.html}), require that submitted work is hosted on GitHub and their review processes make use of GitHub issues.
Peer review of research software by rOpenSci (\url{https://ropensci.org/software-review/}) also makes use of many features of GitHub to streamline and automate parts of the review process.
rOpenSci's efforts have resulted in many well-used R packages for ecology research including rfishbase\textsuperscript{\protect\hyperlink{ref-Xsdcv6q}{33}} and taxize\textsuperscript{\protect\hyperlink{ref-FVBWKkZu}{34}}.
Although the use of GitHub in peer-review is currently restricted to these specialized cases, they may provide a model for peer-review in more traditional EEB journals for manuscripts that make use of code.

@@

The development of research software continues, and so does the need to consider project continuity, especially in EEB where graduate students, research assistants, and postdoctoral fellows often hold relatively short-term positions\textsuperscript{\protect\hyperlink{ref-D4C4k4ak}{37}}.
Often with these types of short-term research projects, once the contract expires, the research software upkeep tends to fall off as the researchers move on to new projects.
Additionally, if the research software is kept on only the researcher's personal devices, it becomes increasingly difficult to access the software and code for future uses.
As projects or contracts end, there should be a handover period of this software in order for the next cohort of researchers to reuse what was already developed\textsuperscript{\protect\hyperlink{ref-4ny1onB0}{8},\protect\hyperlink{ref-D4C4k4ak}{37}}.

GitHub facilitates project continuity among research software and research code by providing tools that make this handover period easier.
As we have already mentioned, using Git for code in Ecology and Evolution can allow for a ``paper trail'' of sorts to be created for the research software, thus allowing for future users of the code access to the entire history of the project\textsuperscript{\protect\hyperlink{ref-3DKwn1sY}{14}}.
Additionally, GitHub allows for repositories and organizations to have designated Code Owners\textsuperscript{\protect\hyperlink{ref-s91uGRZ2}{38}}; these code owners can change through time allowing for the transition of research software from one cohort of researchers to the next\textsuperscript{\protect\hyperlink{ref-iIEKCTLU}{26}}.
There are also multiple means by which to archive software dependency information with GitHub.
Dependencies can be described in README files, added to environment.yml and/or an environment.txt file, or stored in a repository using tools such as renv (\url{https://rstudio.github.io/renv/articles/renv.html}).
This ensures that as packages continue to develop with new syntax, code will still run as it did when it was developed.

Within EEB projects, tasks are often divided among contributors taking various roles (see CRediT taxonomy, \url{https://casrai.org/credit/}).
The creation of project repositories is commonly the purview of those involved in the software, formal analysis, and/or visualization components of the project through their roles as code writers.
However, the structural components of a typical GitHub repository and the derived EEB-specific templates can provide functional ways for other collaborators not contributing to code to be engaged in aspects of repository design in a way that improves institutional memory and facilitates project continuity.
These other collaborators can offer many contributions to repository design and development, and their active involvement can both aid authors ability to act as guarantors of the project, and the clarity and reproducibility of the project for future users.
In (Figure \ref{fig:scatterblob}), we highlight several elements of good repository structure, and the various ways that contributors may interact with them.

>>>>

This manuscript
(\href{https://SORTEE-Github-Hackathon.github.io/manuscript/v/f4614eb536313e122f922ad46d0e4f9f246b724a/}{permalink})
was automatically generated
from \href{https://github.com/SORTEE-Github-Hackathon/manuscript/tree/f4614eb536313e122f922ad46d0e4f9f246b724a}{SORTEE-Github-Hackathon/manuscript@f4614eb}
on June 9, 2022.

@@

Most scientists, including ecologists and evolutionary biologists, are increasingly dependent on computational tools in their research\textsuperscript{\protect\hyperlink{ref-fJWFe93e}{1}}.
Researchers write and use software packages or data analysis code (hereafter, code) as part of their scientific workflow to perform a wide-variety of tasks ranging from data management, data analysis, and study replication, to the application and the development of tools for hypothesis testing.
To maintain code for scientific collaboration requires an efficient and well-documented work-flow\textsuperscript{\protect\hyperlink{ref-1Kqna6l2}{2}}.
To facilitate this process, scientists have been increasingly adopting tools from information and systems technology, such as cloud-based services for documentation and version control (\emph{e.g.}, from the Google Suite, with Docs, Sheets, and Drive; the Microsoft Suite, with Word, Excel, and OneDrive; and GitHub;\textsuperscript{\protect\hyperlink{ref-10ghgV3S8}{3}}).
However, most researchers lack exposure to adequate code development practices and are required to dedicate valuable time and effort to self-teach the use of research-facilitating tools, and thus may be limited in their ability to adhere to adequate standards of scientific code quality and maintenance\textsuperscript{\protect\hyperlink{ref-fJWFe93e}{1},\protect\hyperlink{ref-10SpoByIw}{4},\protect\hyperlink{ref-O6UbstGG}{5}}.
Here, we review and discuss one of the most used web-based platforms for computational version control and collaboration, GitHub, and provide researchers in ecology and evolutionary biology (EEB) with practical workflows aimed at facilitating their scientific code and management process.

With over 73 million registered users, as of 2022, GitHub is the most widely-used web platform for collaborating on computer code\textsuperscript{\protect\hyperlink{ref-nwCtHDCn}{6}}.
GitHub provides a simplified but powerful web interface that allows users to participate in projects, contribute code, report and discuss code bugs, discover existing code and data, and publish new code.
Through version control, users have a detailed, chronological record of the files and directories stored in their repositories (see \protect\hyperlink{definitions}{Box 1};\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7}}).
This workflow provides a strong and clear advantage over receiving, processing and sending files back-and-forth (e.g.~via email), a process that can easily become challenging and time-consuming in projects extending in time and in the number of collaborators\textsuperscript{\protect\hyperlink{ref-4ny1onB0}{8}}.
Through the combination of version control management and the network- and collaboration-based features, GitHub can broadly facilitate openly available source code alongside concomitant collaborative development\textsuperscript{\protect\hyperlink{ref-kEX5dgzK}{9}}.

Git is the version control system that enables all the collaborative tools available on GitHub.
Although the understanding of basic concepts of Git (such as commit, push, pull, checkout; see \protect\hyperlink{definitions}{Box 1}) is necessary, the GitHub web-based platform and its integrated development environments (such as the GitHub Desktop) allow users to perform most repository and data management operations without opening Git command-line sessions.
Still, GitHub users are encouraged to explore and improve their proficiency in Git when feeling comfortable, so they can feel confident in performing more flexible, complex operations (such as integrating changes from one branch into another branch with \texttt{git\ merge} and \texttt{git\ rebase}). Examples of extensive explanations on Git can be found in journal articles\textsuperscript{\protect\hyperlink{ref-kEX5dgzK}{9},\protect\hyperlink{ref-PlcxShQU}{10}}, video tutorials\textsuperscript{\protect\hyperlink{ref-NIS0JOW0}{11}}, and books\textsuperscript{\protect\hyperlink{ref-13jOlVcpp}{12}}.

The expansive GitHub user-community and the numerous resources on how to use GitHub have boosted its growing popularity\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7},\protect\hyperlink{ref-kEX5dgzK}{9},\protect\hyperlink{ref-13jOlVcpp}{12}/,\protect\hyperlink{ref-u5aEVE4B}{13}}.
Nevertheless, although multiple articles have encouraged researchers in EEB to adopt GitHub as part of their research process\textsuperscript{\protect\hyperlink{ref-10ghgV3S8}{3},\protect\hyperlink{ref-3DKwn1sY}{14}}, its use is still not widespread.
First-time users without formal training in information technology may face steep learning curves because GitHub and its features have been centered on collaboration for code development in information systems\textsuperscript{\protect\hyperlink{ref-139b0pSGc}{15}}.
Moreover, domain-specific perspectives and resources providing tractable examples and practical guidance for researchers in EEB on GitHub are scarce (but see https://ourcodingclub.github.io; https://www.openscapes.org).
A common adoption of GitHub for collaborating on a variety of research tasks can ultimately enable EEB researchers to spend less time on creating novel processes for collaboration and more time on their research\textsuperscript{\protect\hyperlink{ref-ydrk01SR}{16}}.
More importantly, increasing the availability of data and code management standards -- of which GitHub is one increasingly important component -- make research more reproducible and collaborative\textsuperscript{\protect\hyperlink{ref-13QX8XU3J}{17}}.

@@

\textbf{release}: At any point a release can be made on GitHub to mark a significant milestone in the progression of a \textbf{repository}. While this GitHub feature is designed with releases of new versions of code in mind (\emph{e.g.}, v1.0.0), it can also be used to create a snapshot of a repository at significant stages like submission, revision, and acceptance of an associated manuscript. \\

@@

Peer review plays a critical role in the scientific evaluation of EEB research.
GitHub provides an open and transparent platform that can be used for either directly providing feedback on research products or building a to-do list from reviewer comments.
Peer review of research code by the open software development community rOpenSci (\url{https://ropensci.org/software-review/}) and of research software and associated manuscripts by the Journal of Open Source Software (\url{https://joss.readthedocs.io/en/latest/submitting.html}) requires that submitted work is hosted on GitHub and their review processes make use of GitHub issues.
rOpenSci's efforts have resulted in many well-used R packages for ecology research including rfishbase\textsuperscript{\protect\hyperlink{ref-Xsdcv6q}{33}} and taxize\textsuperscript{\protect\hyperlink{ref-FVBWKkZu}{34}}.

@@

@@

The development of research code continues, and so does the need to consider project continuity, especially in EEB where graduate students, research assistants, and postdoctoral fellows often hold relatively short-term positions\textsuperscript{\protect\hyperlink{ref-D4C4k4ak}{37}}.
Often with these types of short-term research projects, once the contract expires, the research code upkeep tends to fall off as the researchers move on to new projects.
Additionally, if the code is kept on only the researcher's personal devices, it becomes increasingly difficult to access the code for future uses.
As projects or contracts end, there should be a handover period of this code in order for the next cohort of researchers to reuse what was already developed\textsuperscript{\protect\hyperlink{ref-4ny1onB0}{8},\protect\hyperlink{ref-D4C4k4ak}{37}}.

GitHub facilitates project continuity among research code by providing tools that make this handover period easier.
As we have already mentioned, using Git for code in Ecology and Evolution can allow for a ``paper trail'' of sorts to be created for the code, thus allowing for future users of the code access to the entire history of the project\textsuperscript{\protect\hyperlink{ref-3DKwn1sY}{14}}.
Additionally, GitHub allows for repositories and organizations to have designated Code Owners\textsuperscript{\protect\hyperlink{ref-s91uGRZ2}{38}}; these code owners can change through time allowing for the transition of code from one cohort of researchers to the next\textsuperscript{\protect\hyperlink{ref-iIEKCTLU}{26}}.
There are also multiple means by which to archive code dependency information with GitHub.
Dependencies can be described in README files, added to environment.yml and/or an environment.txt file, or stored in a repository using tools such as renv (\url{https://rstudio.github.io/renv/articles/renv.html}).
This ensures that as packages continue to develop with new syntax, code will still run as it did when it was developed.

Within EEB projects, tasks are often divided among contributors taking various roles (see CRediT taxonomy, \url{https://casrai.org/credit/}).
The creation of project repositories is commonly the purview of those involved in the code, formal analysis, and/or visualization components of the project through their roles as code writers.
However, the structural components of a typical GitHub repository and the derived EEB-specific templates can provide functional ways for other collaborators not contributing to code to be engaged in aspects of repository design in a way that improves institutional memory and facilitates project continuity.
These other collaborators can offer many contributions to repository design and development, and their active involvement can both aid authors ability to act as guarantors of the project, and the clarity and reproducibility of the project for future users.
In (Figure \ref{fig:scatterblob}), we highlight several elements of good repository structure, and the various ways that contributors may interact with them.
